Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-05-08 appendix.tex
typos, redundancies, and moved example earlier

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
For the binary symmetric model $\alphabet=\{0,1\}$.
Define the complement of $\alignmentColumn$ as $\overline{\alignmentColumn}$, and let $\alignmentColumn_{i,i'}$ be the label of the $i'$th tip in the $i$th alignment column.
We define the site split $\siteSplit$ for a $\alignmentColumn_i$ as the set of tips labeled with $1$ in $\alignmentColumn_i$ if the $\nSiteRows$th tip is not labeled with $1$, and as the set of tips labeled with $1$ in $\overline{\alignmentColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$.
Taking such a complement simplifies but does not change the result of likelihood computation because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model.
We define an ``ancestral state split'' in a similar way, where the ancestral state split $\ancestralSplit$ for an internal node $\ancestralStateColumn_i$ is the set of internal nodes labeled with $1$ if the $\nSiteRows$th tip is not labeled with $1$, and as the set of tips labeled with $1$ in $\overline{\ancestralStateColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$
In other words, both site splits and ancestral state splits are defined by whether the $\nSiteRows$th element of $\alignmentColumn_i$ is labeled as $1$.
%EM "ancestral state splits" are an important item in the development but don't get the same level of definition as "site splits". Let's define them here.
%das: done

For a fixed topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization $\ancestralStateColumn_i$.
In this case we cannot use the same complement trick as before: the probability of observing an ancestral state split conditional on a site split is not invariant to taking its complement.

@@

We first fix notation.
\begin{definition}
Let the mapping from site patterns to site splits
\[
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet)
\]
be
\[
\patternToSplit(\alignmentColumn) =
\left\{
    \begin{array}{ll}
        \{i'\in\{1,\ldots,\nSiteRows\}: \alignmentColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}\neq 1,\\
        \{i'\in\{1,\ldots,\nSiteRows\}: \overline{\alignmentColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1,
    \end{array}
\right.
\]
and the mapping from ancestral states and tip states to ancestral state splits
\[
\ancestralToSplit:\alphabet^\nSiteRows\times\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet)
\]
be
\[
\ancestralToSplit(\alignmentColumn, \ancestralStateColumn) =
\left\{
    \begin{array}{ll}
        \{i'\in\{1,\ldots,\nAncestralStateRows\}: \ancestralStateColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}\neq 1,\\
        \{i'\in\{1,\ldots,\nAncestralStateRows\}: \overline{\ancestralStateColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1.
    \end{array}
\right.
\]
Then, given a site pattern--valued random variable $\alignmentColumnRV$, define the random variable
\[
\siteSplitRV := \patternToSplit(\alignmentColumnRV)
\]
and
\[
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
\]
for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$.
\end{definition}
The mapping $\patternToSplit$ operates by returning the tips labeled as $1$ in a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if the set of tips labeled $1$ is not in $\mathcal{P}(\siteSplitSet)$.
The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if the set of tips labeled $1$ in $\alignmentColumn$ is in $\siteSplitSet$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the set of tips labeled $1$ in $\ancestralStateColumn$; otherwise, the set of tips labeled $1$ in $\overline{\alignmentColumn}$ necessarily is in $\siteSplitSet$.

For the $i$th factor of \eqref{eq:full_likelihood},
\[
\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
\]
As a consequence of assuming a binary symmetric model, for some $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ the mapping $\patternToSplit(\alignmentColumn_i)$ has the property
\begin{align*}
    \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \cup \overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) + \Pr(\overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= 2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t)
\end{align*}
where $\overline{Y}$ is the complement of the site-patterned random variable $Y$ and so has the same distribution as $Y$.
Note
\begin{align*}
    \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Thus given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral state splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
%EM Wait, does this establish surjectivity? We need to be able to get all values of P(H).
%das: ah crap, I don't know why I wrote the above. I changed a bit of the notation (sorry!) but I hope this will be more clear?
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it can map values from $\alphabet^\nAncestralStateRows$ to all elements in $\mathcal{P}(\ancestralSplitSet)$.
This can be seen by using the definition of $\ancestralToSplit(\alignmentColumn_i, \cdot)$ and assuming $\alignmentColumn_{i,i'}\neq 1$, where in this case each of the $2^\nAncestralStateRows$ values of $\ancestralStateColumn$ correspond to each of the $2^\nAncestralStateRows$ elements of $\mathcal{P}(\{1,\ldots,\nAncestralStateRows\})$.
The same can be done for the case of $\alignmentColumn_{i,i'}=1$, implying $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective.
From this we have
\begin{align*}
    \max_{\ancestralStateColumn_i} \ \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \max_{\ancestralStateColumn_i} \ \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t)\\
    &= \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)\\
    &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)
\end{align*}
for some $j$ depending on $i$.
Thus, each term in the likelihood can be collapsed into terms relating only to site splits, indexed by $j$,  as opposed to each individual observation, indexed by $i$.

@@

Thus the likelihood in \eqref{eq:profile_likelihood} can be written as a product over site patterns as opposed to sites:
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
for some $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ with $1 \le j \le q$ where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.

@@

We state the Hadamard representation of site split generating probabilities---that is, probabilities of obtaining particular site splits given a tree---following Section 8.6 of \citet{Semple2003-em}.
For each edge $e$ define the edge ``fidelity'' for that edge as
\[
\theta(e) = 1-2p(e)
\]
where $p(e)$ is the probability of a character change along edge $e$.
For an even-sized subset of $Y\subseteq\mathcal{S}$, let the path set $P(Y)$ be the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
For notational convenience, let
\[
p_{\siteSplit_j} := \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t),
\]
for any site split $\siteSplit_j$.
Table~\ref{tab:gen-sitepatprob} contains calculations of site pattern probabilities for the InvFels topology with arbitrary branch lengths and Table~\ref{tab:sitepatprob} site pattern probabilities for the InvFels topology in Fig.~\ref{fig:farris-fels-top}a.

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\})$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1 + x_1x_2 +  y_1y_2 +  w[  x_1 + x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1\}$     & $p_{\{1\}}$       &$1 - x_1x_2 +  y_1y_2 +  w[- x_1 + x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{2\}$     & $p_{\{2\}}$       &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][- y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{3\}$     & $p_{\{3\}}$       &$1 - x_1x_2 +  y_1y_2 +  w[  x_1 - x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{1,2\}$   & $p_{\{1,2\}}$     &$1 - x_1x_2 -  y_1y_2 +  w[- x_1 + x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1,3\}$   & $p_{\{1,3\}}$     &$1 + x_1x_2 +  y_1y_2 +  w[- x_1 - x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{2,3\}$   & $p_{\{2,3\}}$     &$1 - x_1x_2 -  y_1y_2 +  w[  x_1 - x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1,2,3\}$ & $p_{\{1,2,3\}}$   &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][  y_1 - y_2] - x_1y_1x_2y_2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x_1,y_1,x_2,y_2,w\}$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:gen-sitepatprob}
\end{table}

>>>>

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, we describe a given $\alignmentColumn_i$ as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
For the binary symmetric model $\alphabet=\{0,1\}$.
Define the complement of $\alignmentColumn$ as $\overline{\alignmentColumn}$, and let $\alignmentColumn_{i,i'}$ be the label of the $i'$th tip in the $i$th alignment column.
We define the site split $\siteSplit$ for a $\alignmentColumn_i$ as the set of tips labeled with $1$ in $\alignmentColumn_i$ if the $\nSiteRows$th tip is not labeled with $1$, and as the set of tips labeled with $1$ in $\overline{\alignmentColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$.
Taking such a complement simplifies but does not change the result of likelihood computation because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model.

For a fixed topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization $\ancestralStateColumn_i$.
In this case we cannot use the same complement trick as before: the probability of observing an ancestral state split conditional on a site split is not invariant to taking its complement.
We define an ``ancestral state split'' in a similar way to a site split, where the ancestral state split $\ancestralSplit$ for an internal node $\ancestralStateColumn_i$ is the set of internal nodes labeled with $1$ if the $\nSiteRows$th tip---\emph{not} internal node---is not labeled with $1$, and as the set of internal nodes labeled with $1$ in $\overline{\ancestralStateColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$.
In other words, both site splits and ancestral state splits are defined by whether the $\nSiteRows$th element of $\alignmentColumn_i$ is labeled as $1$.
%EM "ancestral state splits" are an important item in the development but don't get the same level of definition as "site splits". Let's define them here.
%das: done

@@

We first fix notation.
\begin{definition}
Let the mapping from site patterns to site splits
\[
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet)
\]
be
\[
\patternToSplit(\alignmentColumn) =
\left\{
    \begin{array}{ll}
        \{i'\in\{1,\ldots,\nSiteRows-1\}: \alignmentColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}\neq 1,\\
        \{i'\in\{1,\ldots,\nSiteRows-1\}: \overline{\alignmentColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1,
    \end{array}
\right.
\]
and the mapping from ancestral states and tip states to ancestral state splits
\[
\ancestralToSplit:\alphabet^\nSiteRows\times\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet)
\]
be
\[
\ancestralToSplit(\alignmentColumn, \ancestralStateColumn) =
\left\{
    \begin{array}{ll}
        \{i'\in\{1,\ldots,\nAncestralStateRows\}: \ancestralStateColumn_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}\neq 1,\\
        \{i'\in\{1,\ldots,\nAncestralStateRows\}: \overline{\ancestralStateColumn}_{i,i'}=1\}  & \mbox{if } \alignmentColumn_{i,\nSiteRows}=1.
    \end{array}
\right.
\]
Then, given a site pattern--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$, define the random variables
\[
\siteSplitRV := \patternToSplit(\alignmentColumnRV)
\]
and
\[
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV).
\]
\end{definition}
The mapping $\patternToSplit$ operates by returning the tips labeled as $1$ in a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if the set of tips labeled $1$ is not in $\mathcal{P}(\siteSplitSet)$.
The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if the set of tips labeled $1$ in $\alignmentColumn$ is in $\mathcal{P}(\siteSplitSet)$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the set of tips labeled $1$ in $\ancestralStateColumn$; otherwise, the set of tips labeled $1$ in $\overline{\alignmentColumn}$ necessarily is in $\mathcal{P}(\siteSplitSet)$.

For the $i$th factor of \eqref{eq:full_likelihood},
\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
As a consequence of assuming a binary symmetric model, for some $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ the mapping $\patternToSplit(\alignmentColumn_i)$ has the property
    \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \cup \overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) + \Pr(\overline{\alignmentColumnRV}=\alignmentColumn_i \mid \tau, t)\\
    &= 2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t)
where $\overline{Y}$ is the complement of the site pattern--valued random variable $Y$ and has the same distribution as $Y$.
Note
    \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
Thus given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subseteq\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral state splits for that particular site split, topology and set of branch lengths, i.e., $\ancestralSplitPartition_j(\tau, t)$ is a set of sets of most likely internal node labels.
Here, $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
%EM Wait, does this establish surjectivity? We need to be able to get all values of P(H).
%das: ah crap, I don't know why I wrote the above. I changed a bit of the notation (sorry!) but I hope this will be more clear?
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it can map values from $\alphabet^\nAncestralStateRows$ to all elements in $\mathcal{P}(\ancestralSplitSet)$.
This can be seen by using the definition of $\ancestralToSplit(\alignmentColumn_i, \cdot)$ and assuming $\alignmentColumn_{i,\nSiteRows}\neq 1$, where in this case each of the $2^\nAncestralStateRows$ values of $\ancestralStateColumn$ correspond to each of the $2^\nAncestralStateRows$ elements of $\mathcal{P}(\{1,\ldots,\nAncestralStateRows\})$.
The same can be done for the case of $\alignmentColumn_{i,\nSiteRows}=1$, implying $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective.
From this we have
    \max_{\ancestralStateColumn_i} \ \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \max_{\ancestralStateColumn_i} \ \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t)\\
    &= \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)\\
    &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)
for some $j$.
Thus, each term in the likelihood can be collapsed into terms relating only to site splits and ancestral state splits, indexed by $j$, as opposed to individual observations, indexed by $i$.

\subsection*{Example}
We follow with an expository example computing these probabilities and likelihoods.
Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
\[
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
\]
Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
\[
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
\]
in general.

@@

The likelihood in \eqref{eq:profile_likelihood} can be written as
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
for $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ and some $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ with $1 \le j \le q$ where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.

@@

We state the Hadamard representation of site split generating probabilities---that is, probabilities of obtaining particular site splits given a tree---following Section 8.6 of \citet{Semple2003-em}.
For each edge $e$ define the edge ``fidelity'' for that edge as
\[
\theta(e) = 1-2p(e)
\]
where $p(e)$ is the probability of a character change along edge $e$.
For an even-sized subset of $Y\subseteq\mathcal{S}$, let the path set $P(Y)$ be the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
For notational convenience, let
\[
p_{\siteSplit_j} := \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t),
\]
for any site split $\siteSplit_j$.
Table~\ref{tab:gen-sitepatprob} contains calculations of site split probabilities for the InvFels topology with arbitrary branch lengths and Table~\ref{tab:sitepatprob} site split probabilities for the InvFels topology in Fig.~\ref{fig:farris-fels-top}a.

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\})$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1 + x_1x_2 +  y_1y_2 +  w[  x_1 + x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1\}$     & $p_{\{1\}}$       &$1 - x_1x_2 +  y_1y_2 +  w[- x_1 + x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{2\}$     & $p_{\{2\}}$       &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][- y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{3\}$     & $p_{\{3\}}$       &$1 - x_1x_2 +  y_1y_2 +  w[  x_1 - x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{1,2\}$   & $p_{\{1,2\}}$     &$1 - x_1x_2 -  y_1y_2 +  w[- x_1 + x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1,3\}$   & $p_{\{1,3\}}$     &$1 + x_1x_2 +  y_1y_2 +  w[- x_1 - x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{2,3\}$   & $p_{\{2,3\}}$     &$1 - x_1x_2 -  y_1y_2 +  w[  x_1 - x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1,2,3\}$ & $p_{\{1,2,3\}}$   &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][  y_1 - y_2] - x_1y_1x_2y_2$\\
    \hline
\end{tabular}
\caption{Site split probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x_1,y_1,x_2,y_2,w\}$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:gen-sitepatprob}
\end{table}
